### 2024-07-01 causal-inference.qmd
20240701 - bias

>>>>

### Confounding {#sec-causalDiagramConfounding}

Confounding occurs when two variables have a spurious or noncausal association because both variables are caused by another variable(s) [@DOnofrio2020].
That is, two variables share a common cause, and the common cause leads the variables to be associated even though they are not causally related.
The common cause—i.e., the variable that influences the two variables—is known as a confound (or confounder).

### Mediation {#sec-causalDiagramMediation}

### Collider Bias {#sec-colliderBias}

Collider bias occurs when two variables influence a third variable, the collider [@DOnofrio2020].
That is, a collider is a variable that is caused by two other variables.

#### M-Bias {#sec-causalDiagramMBias}

Collider bias may also occur when neither variable of interest is a direct cause of the collider [@Lederer2019].
M-bias is a form of collider bias that occurs when two variables that are not causally related, `A` and `B`, both influence a collider, `C`, and each (`A` and `B`) also influences a separate variable—e.g., `A` influences `D` and `B` influences `E`.
M-bias is so-named from the M-shape of the DAG.

### Butterfly Bias {#sec-colliderBias}

Butterfly bias occurs when both confounding and M-bias are present.
Butterfly bias is so-named from the butterfly shape of the DAG.

### Selection Bias {#sec-causalDiagramSelectionBias}

Selection bias occurs when the selection of participants or their inclusion in analyses depends on the variables being studied.
